Title Goes Here \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2021-03-04 analysis/paper/paper.tex
Draft resultados

We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between 70°S and 45°S for each spring and computing the Fourier spectrum.

@@

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. (arbitrary units). The numbers at the top-left of each panel is the variance explained by each EOF.}
\end{figure}

@@

Another characteristic seen in Figure \ref{fig:eof-naive} is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa. Furthermore, Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between temporal series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.

@@

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S. This also nearly minimises the correlation with the imaginary part (\ref{fig:rotations-pc1}).

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (\ref{fig:rotations-pc2}). It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase. If the relationship is very weak, the rotation cannot generate a strong relationship.

@@

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Leading 2 EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Spatial patterns. Real part in shading, imaginary part in contours}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-2-1.pdf}
\caption{\label{fig:ceofs-2}Temproal series.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/extended-series-1.pdf}
\caption{\label{fig:extended-series}Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.}
\end{figure}

Figure \ref{fig:extended-series} shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the ``hybrid ERA5'' reanalysis). There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \ref{fig:extended-series}a.1, p-value = 0.0011). There is no significant trend in any of the complex parts of EOF2.

@@

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression of EOF1 with mean air temperature between 80ºS and 25ºS.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/o3-regr-1.pdf}
\caption{\label{fig:o3-regr}Regression of standarised EOF1 and Total Ozone column.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/wave1-o3-1.pdf}
\caption{\label{fig:wave1-o3}EOF1 and wave1 o3 amplitude.}
\end{figure}

Indeed, Figure \ref{fig:o3-regr} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1. The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.87 (CI: 0.76 -- 0.93). As expected from the location of anomalies in Figure @ref(fig: o3-regr), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).

@@

\begin{figure}
\centering
\includegraphics{../figures/psa-eof2-1.pdf}
\caption{\label{fig:psa-eof2}EFO2 and PSA}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/phase-histogram-1.pdf}
\caption{\label{fig:phase-histogram}Histogram of phase distributions of EOF2.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/fake-eof-1.pdf}
\caption{\label{fig:fake-eof}Principal compomnents of a sinthetic data set of geopotential height as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like annomalies with random amplitude.}
\end{figure}

@@

To better explore the relationship between tropical forcing and phase of the EOF2, Figure \ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50\% of the years. In years with positive ENSO, the phase of the EOF2 is always around +90º (corresponding with positive imaginary part) and vice versa. In years with near neutral ENSO, the phase of the EOF2 is much more variable. The black line in Figure \ref{fig:enso-phase} tries to quantify this relationship. Is the equation \(ONI = 1.1\sin{phase}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight. The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001.

Figure \ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years. The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is 0.44 (CI: 0.16 -- 0.66). This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest EOF2 magnitude. If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)). Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value = = 0.2). Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.

@@

Figure \ref{fig:pp-america} shows regression maps of seasonal precipitation with each EOF in South America. EOF1 (Figure \ref{fig:pp-america} column a) is associated with increased precipitation over Paraguay and and some parts of Brazil. Although these positive anomalies are not very strong. The strongest precipitation anomalies are the ones associated with the Imaginary EOF2. The positive anomalies on Southeaster South America (SESA) and Chile, and negative anomalies over Southern Brazil is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it's also virtually identical to the precipitation anomalies associated with the Asymmetric SAM \citep{campitelli2021}. This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously. The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA. The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure \ref{fig:enso-phase} (not shown).

\begin{figure}
\centering
\includegraphics{../figures/pp-africa-1.pdf}
\caption{\label{fig:pp-africa}Same as Figure \ref{fig:pp-america} but for South of Africa.}
\end{figure}

@@

@@

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\hypertarget{chosen-rotations-of-the-eofs}{%
\subsection{Chosen rotations of the EOFs}\label{chosen-rotations-of-the-eofs}}

\begin{figure}
\centering
\includegraphics{../figures/rotations-pc1-1.pdf}
\caption{\label{fig:rotations-pc1}Rotations of PC1}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/rotations-pc2-1.pdf}
\caption{\label{fig:rotations-pc2}R2 between EOF2 Real and Imaginary parts and ONI index for different rotation parameters of the real and imaginary parts.}
\end{figure}

>>>>

We compute the amplitude and phase of the zonal wave 1 by averaging SON TOC between 70°S and 45°S for each Spring and computing the Fourier spectrum.

@@

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units). The numbers at the top-left of each panel ate the variance explained by each EOF.}
\end{figure}

@@

Another characteristic seen in Figure \ref{fig:eof-naive} is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa. Furthermore, Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.

@@

\begin{figure}
\centering
\includegraphics{../figures/rotations-1.pdf}
\caption{\label{fig:rotations}a) Correlation of the Real and Imaginary EOF1 with wave 1 amplitude of Total Ozone Column averaged between 70°S and 45°S and b) \(R^2\) between the Real and Imaginary EOF2 and the ONI index for different values of rotation. The vertical line marks the selected rotation, which maximises de correlation.}
\end{figure}

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S. This also nearly minimises the correlation with the imaginary part (Figure \ref{fig:rotations}a).

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure \ref{fig:rotations}b). It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase. If the relationship is very weak, the rotation cannot generate a strong relationship.

@@

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Real part in shading, imaginary part in contours}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-2-1.pdf}
\caption{\label{fig:ceofs-2}Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/extended-series-1.pdf}
\caption{\label{fig:extended-series}Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern and scaled to zero mean and unit standard deviation.}
\end{figure}

Figure \ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the ``hybrid ERA5'' reanalysis). There is a downward trend in the real part of EOF1 in the hybrid reanalysis (Figure \ref{fig:extended-series}a.1, p-value \textless{} 0.001). There is no significant trend in any of the complex parts of EOF2.

@@

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression of EOF1 with mean air temperature averaged between 80ºS and 25ºS for the period 1979 -- 2019.}
\end{figure}

@@

(raf:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column for the 1979 -- 2019 period.

\begin{figure}
\centering
\includegraphics{../figures/o3-regr-1.pdf}
\caption{\label{fig:o3-regr}(raf:o3-regr-cap)}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/wave1-o3-1.pdf}
\caption{\label{fig:wave1-o3}Relationship between EOF1 amplitude and phase with amplitude and phase of the zonal wave 1 in Total Ozone Column averaged between 70°S and 45°S for the period 1979 -- 2019.}
\end{figure}

Indeed, Figure \ref{fig:o3-regr} shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1. The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is 0.86 (CI: 0.76 -- 0.93). As expected from the location of anomalies in Figure @ref(fig: o3-regr), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).

@@

\begin{figure}
\centering
\includegraphics{../figures/psa-eof2-1.pdf}
\caption{\label{fig:psa-eof2}Relationship between the Real and Imaginary parts of EOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies \citep{mo2001}. For the period 1979 -- 2019.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/phase-histogram-1.pdf}
\caption{\label{fig:phase-histogram}Histogram showing phase distribution of EOF2 for the period 1979 -- 2019.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{../figures/fake-eof-1.pdf}
\caption{\label{fig:fake-eof}EOFs of a synthetic data set of 41 times steps of geopotential height fields as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like anomalies with random amplitude.}
\end{figure}

@@

To better explore the relationship between tropical forcing and phase of the EOF2, Figure \ref{fig:enso-phase} plots the ONI index and the phase of the EOF2 for each year between 1979 and 2019, highlighting years in which the magnitude of EOF2 was higher than 50\% of the years. In years with positive ENSO, the phase of the EOF2 is always around +90º (corresponding with positive imaginary part) and vice versa. In years with near neutral ENSO, the phase of the EOF2 is much more variable. The black line in Figure \ref{fig:enso-phase} tries to quantify this relationship. Is the equation \(ONI = 1.1\sin{(phase)}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight. The \(r^2\) corresponding to the fit is 0.56, with p-value \textless{} 0.001.

Figure \ref{fig:enso-phase} suggest that strong EOF2 years tend to coincide with strong ENSO years. The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is 0.44 (CI: 0.16 -- 0.66). This relationship, however, appears to be driven only by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which also coincide with the three years with strongest EOF2 magnitude. If those years are removed, the correlation becomes non-significant (0.044 (CI: -0.28 -- 0.36)). Furthermore, even when using all the datapoints, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value = 0.2). Therefore, the relationship between the magnitude of the EOF2 train wave and ENSO remains uncertain.

@@

Figure \ref{fig:pp-america} shows regression maps of seasonal precipitation with each EOF in South America. EOF1 is associated with slight decreases in precipitation in Southern Brazil and Paraguay, although these are not statistically significant \ref{fig:pp-america} column a. The strongest precipitation anomalies are the ones associated with the Imaginary EOF2. The positive anomalies on Southeaster South America (SESA) and Chile, and negative anomalies over Southern Brazil is a well known springtime precipitation signature of ENSO \citep{cai2020a} and it's also virtually identical to the precipitation anomalies associated with the Asymmetric SAM \citep{campitelli2021}. This is not surprising considering the close relationship between the ONI, the Asymmetric SAM index and the Imaginary EOF2 shown previously, but further consolidates the identification of this mode with the PSA pattern. The Real EOF2, on the other hand, is associated with negative precipitation anomalies in a smaller area of SESA. The relationship between precipitation anomalies in SESA and the phase of EOF2 follows a curve similar to that of Figure \ref{fig:enso-phase} (not shown).

@@

On the antipodes, precipitation anomalies in New Zealand and neighbouring islands associated with EOF1 are also weak and not statistically significant (Figure \ref{fig:pp-oceania} column a) whereas EFO2 is associated with large and expansive anomalies. The Real EOF2 is associated with positive seasonal anomalies in Northern and Eastern Australia, and negative ones over New Zealand, while precipitation anomalies associated with Imaginary EOF2 are negative over all Eastern Australia. Again, these negative anomalies are similar to the springtime precipitation anomalies observed in relation with the Asymmetric SAM \citep{campitelli2021}.

\begin{figure}
\centering
\includegraphics{../figures/australia-pp-phase-1.pdf}
\caption{\label{fig:australia-pp-phase}Same as Figure \ref{fig:enso-phase} but for precipitacion anomalies averaged between 130°E and 155°E and 40°S and 15°S (box shown in Figure \ref{fig:pp-oceania} column b).}
\end{figure}

To better understand the relationship between precipitation anomalies over Western Australia and the EOF2, Figure \ref{fig:australia-pp-phase} plots precipitation anomalies averaged over Eastern Australia (box in Figure \ref{fig:pp-oceania}) as a function of EOF2 phase. The black line is the equation \(\mathrm{Precipitation} = -0.26\sin{(\mathrm{phase})} + 0.22\cos{(\mathrm{phase})}\), whose coefficient we fitted by weighted least squares using the magnitude of the EFO2 as weight. The \(r^2\) corresponding to the fit is 0.37, with p-value \textless{} 0.001. There is a strong relationship between the phase of the EOF2 and precipitation anomalies in Eastern Australia, but Real and Imaginary phases we chose are not aligned with the direction that maximises this relationship. If one were to perform a more in-depth analysis of the relationship between this pattern and precipitation in this region, one would conceivably align one of the axis (either the Real or the Imaginary) to the one that maximises it.

@@
